prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-27 inst/MyArticle/paper.Rmd
new Biocpkg function

```{r setup, include=FALSE}
## See ACS guidelines [https://publish.acs.org/publish/author_guidelines?coden=ancham#manuscript_text_components]
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 5)
ggplot2::theme_set(ggplot2::theme_classic())
library(BiocStyle)
options(width = 70)
```

@@

Several data analysis packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` determines the statistical model from the structure of the sample annotation, which allows users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows the specification of a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package `r BiocStyle::Biocpkg("PECA")` performs a roll-up of peptide level differences and peptide level $p$-value estimates obtained from `r BiocStyle::Biocpkg("limma")` or `r BiocStyle::Biocpkg("PECA")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` combines two models: robust linear models fitted to protein abundances and a quasibinomial generalized linear model fitted to peptide counts, into Hurdle model. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

Of note are the various approaches to handle missing observations, which are inherent in mass spectrometric bottom-up experiments. For instance, `r BiocStyle::Biocpkg("MSstats")` handles missing data by feature filtering and imputation. Other means of modeling missing observations are the Hurdle models discussed by @goeminne2020msqrob, while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

We can use all the R packages discussed when analyzing parallel-group designs using a single explanatory variable and contrasting groups. However, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that, for instance, `r BiocStyle::Biocpkg("limma")` and `r BiocStyle::Biocpkg("proDA")` allow us to fit a comprehensive variety of models and test various hypotheses; however, good knowledge of the design matrix specification using the R formula interface is required [@law2020guide].

@@

Furthermore, `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N [@Grossmann2010], Tukey's median polish  [@tukey1977exploratory], robust linear models [@goeminne2020msqrob]. Peptide or protein abundances can then be scaled and transformed using robust scaling, _quantile_ normalization or `r BiocStyle::Biocpkg("vsn")` to remove systematic differences among samples and heteroscedasticity. In this respect, `r BiocStyle::Githubpkg("fgcz/prolfqua")` can be compared with R packages such as `r BiocStyle::Biocpkg("DEP")` [@DEP2018] or `r BiocStyle::Biocpkg("POMA")` [@POMA2021] which support the entire differential expression analysis pipeline.

We use the Ionstar [@shen2018ionstar] and CPTAC (available from `r BiocStyle::Biocpkg("msdata")`) dataset to benchmark the modeling methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`. Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. The quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and afterward subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion ($FDP$).

@@

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r BiocStyle::Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences among the various input formats in column names and enables the use of the methods without additional arguments. An example code for creating an MSFragger[@yu2020fast] configuration can be found in the Appendix. We implemented methods that transform the data into tidy data for popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features in RStudio [@RStudio2022], and help to implement argument free functions (see Supplementary Figure SX).

@@

R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending; @law2020guide]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. In addition, we implement features specific to high throughput experiments, such as the empirical Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)). We specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform differential expression analysis given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or to impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate differential expression analysis results of external tools such as `r BiocStyle::Biocpkg("proDA")` or of SAINTexpress [@teo2014saintexpress] used to analyze data from protein interaction studies.

@@

The peptide abundances (from the MaxQuant _peptide.txt_ file) were $log_2$ transformed and subsequently scaled, where median and the mean absolute deviation were obtained from the human proteins only. We removed _one hit wonders_, i.e., proteins with a single peptide assignment. Protein abundances are inferred from the peptide intensities using Tukey's median polish. Finally, we fitted the fixed effect linear models, the dropout model `r BiocStyle::Biocpkg("proDA")` to protein abundances, the mixed effect linear model, the `r BiocStyle::Biocpkg("PECA")` model, and hurdle model `r BiocStyle::Biocpkg("msqrob2")` to peptide intensities.

@@

The package `r BiocStyle::Githubpkg("fgcz/prolfqua")` provides functions to determine the vector of $parameter$ weights $c$, from a linear model and a contrast specification string.

The following code illustrates how to specify a model with two factors and an interaction term and define the differences we want to estimate. The example dataset was acquired in two batches (p2370, p2691). In each, we measured Yeast samples grown under two different conditions (Glucose and Ethanol). This example is taken from the `r BiocStyle::Githubpkg("fgcz/prolfqua")` package vignette file. To estimate treatment differences, we first specify the linear model that explains the observed protein abundances (transformedIntensities) using the explanatory variables `condition_` and `batch_` and the interaction term  `condition_:batch_`. Then, after fitting the model, we can estimate the treatment difference among the groups `Ethanol` and `Glucose` defined by the factor `condition` and similarly among the groups defined by `batch_`. Furthermore, we can examine differences between `Ethanol` and `Glucose` within each batch. We also can assess if these differences are the same or are different in both batches, i.e., if there is an interaction between condition and batch. We also show the code to compute the array of weights $c$ used to multiply the model coefficient $\beta$.

```{r contrast2, eval = TRUE, echo = TRUE}

Yeast2Factor <- prolfqua::prolfqua_data("data_Yeast2Factor")
Yeast2Factor$data |> dplyr::select(condition_, batch_) |> dplyr::distinct()
mlm <- prolfqua::strategy_lm(
  transformedIntensity ~ condition_ + batch_ + condition_:batch_
  )
mm <- prolfqua::build_model(Yeast2Factor$data, mlm, subject_Id = "protein_Id")

Contrasts <- c(
"diff_condition" = "condition_Ethanol - condition_Glucose",
"diff_batch" = "batch_p2370 - batch_p2691",
"diff_within_p2370" = "`condition_Glucose:batch_p2370` - `condition_Ethanol:batch_p2691`",
"diff_within_p2691" = "`condition_Glucose:batch_p2691` - `condition_Ethanol:batch_p2691`",
"diff_of_difference" = "diff_within_p2370 - diff_within_p2691")

linfct <- prolfqua::linfct_from_model(mm$modelDF$linear_model[[1]],
                                      as_list = FALSE)
mC <- prolfqua::linfct_matrix_contrasts(linfct, Contrasts)
## see Table 3 for content of mC
```

Table \@ref(tab:contrcoefficient) shows weights $c$ for each of the five contrasts we specified.

```{r contrcoefficient}
knitr::kable(mC, caption = "Weights $c$ for each of the contrasts (rows) which will be applied to the model parameter (columns).",format = "latex")
```

Given the weights $c$ contrasts from model parameters obtained with `r BiocStyle::Biocpkg('proDA')`, or `r BiocStyle::Biocpkg('msqrob2')` can be determined, which will enable us to implement adapters to `proDA` or `msqrob2`.

@@

Where $n_g$ is the number of groups and $n$ is the number of observations.
If variance can not be estimated for a protein, because there are too few observations in other groups, we use the median pooled variance of all other proteins in the dataset.

This method is implemented in the class `ContrastSimpleImpute` (see Figure \@ref(fig:ContrastUML)).

@@

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify the integration of `r BiocStyle::Githubpkg("fgcz/prolfqua")` with Bioconductor based workflows there is a method that converts the `LFQData` class into a `r BiocStyle::Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code where functions can have few arguments. Auto-completion support for R6 classes in the Rstudio editor makes it easy for novices to explore `r BiocStyle::Githubpkg("fgcz/prolfqua")'s functionality (see Supplementary Figure S[X]). 
To ease the usage barriers of the R package to users not familiar with statistics and R programming, we developed an application based on the prolfqua package into our data management platform [B-Fabric](https://fgcz-bfabric.uzh.ch/) [@bfabric,@bfabricjib2022]. The B-Fabric system runs a computing infrastructure controlled by a local resource management system that supports cloud-bursting [@VMMAD]. This integration enables users to select the input data and basic settings in a graphical user interface (GUI).

In this GUI application, we determine the model formula and which comparisons to compute from the sample annotation provided in tabular form, similarly to SAINTexpress and `r BiocStyle::Biocpkg("MSstats")`. However, this application can analyze only parallel group designs with and without repeated measurements or factorial designs without interactions. The user receives a report including input files, the R markdown file, and R scripts necessary to replicate the analysis using their in-house R installation. In this way, `r BiocStyle::Githubpkg("fgcz/prolfqua")` and [B-Fabric](https://fgcz-bfabric.uzh.ch/) and helps scientists to meet requirements from funding agencies, journals, and academic institutions while publishing their data according to the FAIR (Findable, Accessible, Interoperable, and Reusable) [@FAIR] data principles. We are working on creating a shiny stand-alone application with the described functionality and making it available soon.

@@

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes receiver operator curves (ROC) and computes partial areas under those curves ($pAUC$) and other scores, e.g., the false discovery proportion $FDP$. We use those scores, i.e. the $pAUC$ at $10%$ $FDR$ and the $FDP$, to examine how well the methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` model quantitative mass spectrometric high throughput data and compare them with the results produced by `r BiocStyle::Biocpkg("MSstats")`, `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Biocpkg("msqrob2")`. Table \@ref(tab:describeBenchmarkedModels) summarizes all the modelling methods we evaluated.

@@

p1_pAUC <- ggplot2::ggplot(resAllB, ggplot2::aes(x = Name, y = pAUC_10)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::facet_wrap(~whatfix)  + 
  ggplot2::coord_cartesian(ylim = c(min(resAllB$pAUC_10),max(resAllB$pAUC_10))) + 
  ggplot2::theme_minimal() + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5)) +
  ggplot2::geom_hline(ggplot2::aes(yintercept = meanpAUC_10), color = "red") +
  ggplot2::xlab("") +
  ggplot2::labs(tag = "A")

@@

When comparing differential expression analysis performance, a relevant parameter is the number of proteins for which a method estimates differences (see Figure \@ref(fig:figure3) Panel A), which indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4046 proteins with at least two peptides, there are in total $16184$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as the $pAUC$. For instance, we observe that `r BiocStyle::Biocpkg("MSstats")` estimates $16058$ group differences while the mixed effect models estimates the fewest with $15940$. Hence, to conclude that one method shows a better performance, it does not suffice if the $pAUC$ is greater, but the number of proteins with differential expression results needs to be equal or larger.

Figure \@ref(fig:figure3) Panel B shows how various estimates obtained from the models, i.e., the difference between groups, $t$-statistics and scaled $p$-values, allow identifying true positives (TP) given a false positive rate (FPR) of $10%$ by displaying the partial area under the ROC. Ordering proteins by the $t$-statistic or $p$-value leads to a higher $pAUC_{10}$ than when ordering by the estimated difference among groups. Suppose an accurate estimate of the difference among groups is essential. In that case, the linear models fitted to protein intensities, calculated using Tukey's median polish, perform better than those directly modeling peptide intensities, e.g., `ropeca` or `prot_mixed_effect_moderated` (see Table \@ref(tab:describeBenchmarkedModels) Abundance column). We speculate that outliers at the peptide level do not affect the protein estimates when using Tukey's median polish method. This hypothesis is examined in the benchmark, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`.

There are only minor differences in the $pAUC_{10}$ between the $t$-statistics or the scaled $p$-value  (see Figure \@ref(fig:figure3) Panel A). Interestingly, the $pAUC$ increases when using $p$-values instead of the $t$-statistics for linear models, while it decreases for mixed effect models. The reason is an erroneous denominator degree of freedom estimation for many proteins, a known problem in the case of mixed effect models.

We also benchmark if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:figure3) Panel C shows the $FDP$, obtained from the confusion matrix, as a function of the $FDR$ determined from the model. Most lines are below the diagonal, which indicates that the $FDR$ estimates are modestly conservative for this particular benchmark dataset. In the case of `r BiocStyle::Biocpkg("MSstats")`, we observe a high proportion of false discoveries for small $FDR$ values. In the case of the ROPECA method, the $FDR$ estimates, obtained by applying the Benjamini-Hochberg correction to the Beta distribution-based regulation probabilities, strongly overestimate the $FDP$.

@@

nrgg <- dds |> ggplot2::ggplot(ggplot2::aes(x = name, y = nrcontrasts )) + 
  ggplot2::geom_bar(stat = "identity", fill = "white", colour = "black") + 
  ggplot2::coord_cartesian(ylim = c(min(dds$nrcontrasts) - 200, max(dds$nrcontrasts) + 10)) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5)) +
  ggplot2::geom_text(ggplot2::aes(label = paste0(round(nrcontrasts, digits = 1), paste0("  (",round(percent, digits = 1),"%)"))),
            vjust = 0, hjust = -0.2, angle = -90) #+ 
pB <- nrgg + ggplot2::labs(tag = "B")

@@

pC <- ddb |> ggplot2::ggplot(ggplot2::aes(y = FDP,  x  = FDR )) + 
  ggplot2::geom_line(ggplot2::aes(color = model_name, linetype = model_name)) +
  ggplot2::facet_wrap(~contrast) + 
  ggplot2::geom_abline(intercept = 0, slope = 1, color = 2) + 
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(tag = "C") + ggplot2::xlim(c(0,0.5)) + ggplot2::ylim(c(0,0.5))
hlay = rbind(c(2,1,1),
            c(2,3,3))

gridExtra::grid.arrange(p1_pAUC, pB, pC, layout_matrix = hlay)

@@

Using a benchmark dataset whose ground truth is known (see Methods), we assessed the performance of different modeling approaches implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` (Tables \@ref(tab:prolfquaModels) and \@ref(tab:describeBenchmarkedModels)), `r BiocStyle::Biocpkg("MSstats")`, `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Biocpkg("msqrob2")`. Table \@ref(tab:describeBenchmarkedModels) summarizes which methods we have evaluated, which MaxQuant files we used as input, and if the models are fitted to peptide or protein intensities. We make the R-markdown files to replicate the benchmarking available at `r BiocStyle::Githubpkg("wolski/prolfquabenchmark")`.

>>>>

```{r setup, include=FALSE}

Biocpkg <- function(pkg, label = NULL){
  url <- file.path("https://bioconductor.org/packages", pkg)
  BiocStyle:::labelled_link(pkg , label , url)
}

## See ACS guidelines [https://publish.acs.org/publish/author_guidelines?coden=ancham#manuscript_text_components]
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 5)
ggplot2::theme_set(ggplot2::theme_classic())
library(BiocStyle)
options(width = 70)
```

@@

Several data analysis packages exist to model mass spectrometry protein quantification experiments, e.g., `r Biocpkg("limma")` [@Ritchie2015], `r Biocpkg("MSstats")` [@Choi2014], `r Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r Biocpkg("msqrob2")` [@Goeminne2016] or `r Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in R. Each of them has some unique features, for example, `r Biocpkg("MSstats")` determines the statistical model from the structure of the sample annotation, which allows users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r Biocpkg("limma")` allows the specification of a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package `r Biocpkg("PECA")` performs a roll-up of peptide level differences and peptide level $p$-value estimates obtained from `r Biocpkg("limma")` or `r Biocpkg("PECA")`, to protein level estimates. Furthermore, `r Biocpkg("msqrob2")` combines two models: robust linear models fitted to protein abundances and a quasibinomial generalized linear model fitted to peptide counts, into Hurdle model. The `r Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

Of note are the various approaches to handle missing observations, which are inherent in mass spectrometric bottom-up experiments. For instance, `r Biocpkg("MSstats")` handles missing data by feature filtering and imputation. Other means of modeling missing observations are the Hurdle models discussed by @goeminne2020msqrob, while the R package `r Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

We can use all the R packages discussed when analyzing parallel-group designs using a single explanatory variable and contrasting groups. However, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that, for instance, `r Biocpkg("limma")` and `r Biocpkg("proDA")` allow us to fit a comprehensive variety of models and test various hypotheses; however, good knowledge of the design matrix specification using the R formula interface is required [@law2020guide].

@@

Furthermore, `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N [@Grossmann2010], Tukey's median polish  [@tukey1977exploratory], robust linear models [@goeminne2020msqrob]. Peptide or protein abundances can then be scaled and transformed using robust scaling, _quantile_ normalization or `r Biocpkg("vsn")` to remove systematic differences among samples and heteroscedasticity. In this respect, `r BiocStyle::Githubpkg("fgcz/prolfqua")` can be compared with R packages such as `r Biocpkg("DEP")` [@DEP2018] or `r Biocpkg("POMA")` [@POMA2021] which support the entire differential expression analysis pipeline.

We use the Ionstar [@shen2018ionstar] and CPTAC (available from `r Biocpkg("msdata")`) dataset to benchmark the modeling methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r Biocpkg("MSstats")` and `r Biocpkg("proDA")`. Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. The quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and afterward subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion ($FDP$).

@@

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences among the various input formats in column names and enables the use of the methods without additional arguments. An example code for creating an MSFragger[@yu2020fast] configuration can be found in the Appendix. We implemented methods that transform the data into tidy data for popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features in RStudio [@RStudio2022], and help to implement argument free functions (see Supplementary Figure SX).

@@

R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending; @law2020guide]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. In addition, we implement features specific to high throughput experiments, such as the empirical Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)). We specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform differential expression analysis given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or to impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate differential expression analysis results of external tools such as `r Biocpkg("proDA")` or of SAINTexpress [@teo2014saintexpress] used to analyze data from protein interaction studies.

@@

The peptide abundances (from the MaxQuant _peptide.txt_ file) were $log_2$ transformed and subsequently scaled, where median and the mean absolute deviation were obtained from the human proteins only. We removed _one hit wonders_, i.e., proteins with a single peptide assignment. Protein abundances are inferred from the peptide intensities using Tukey's median polish. Finally, we fitted the fixed effect linear models, the dropout model `r Biocpkg("proDA")` to protein abundances, the mixed effect linear model, the `r Biocpkg("PECA")` model, and hurdle model `r Biocpkg("msqrob2")` to peptide intensities.

@@

The package `r BiocStyle::Githubpkg("fgcz/prolfqua")` provides functions to determine the vector of $parameter$ weights $c$, from a linear model and a contrast specification string. In the Supplement we provided an example, how to specify contrasts for a dataset with two explanatory variable with an interaction term (Section "Specifying Contrasts for Models with two Factors and Interaction Term").

@@

Where $n_g$ is the number of groups and $n$ is the number of observations.
If variance can not be estimated for a protein, because there are too few observations in other groups, we use the median pooled variance of all other proteins in the dataset. This method is implemented in the class `ContrastSimpleImpute` (see Figure \@ref(fig:ContrastUML)).

@@

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify the integration of `r BiocStyle::Githubpkg("fgcz/prolfqua")` with Bioconductor based workflows there is a method that converts the `LFQData` class into a `r Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code where functions can have few arguments. Auto-completion support for R6 classes in the Rstudio editor makes it easy for novices to explore `r BiocStyle::Githubpkg("fgcz/prolfqua")`'s functionality (see Supplementary Figure S[X]). 
To ease the usage barriers of the R package to users not familiar with statistics and R programming, we developed an application based on the prolfqua package into our data management platform [B-Fabric](https://fgcz-bfabric.uzh.ch/) [@bfabric,@bfabricjib2022]. The B-Fabric system runs a computing infrastructure controlled by a local resource management system that supports cloud-bursting [@VMMAD]. This integration enables users to select the input data and basic settings in a graphical user interface (GUI).

In this GUI application, we determine the model formula and which comparisons to compute from the sample annotation provided in tabular form, similarly to SAINTexpress and `r Biocpkg("MSstats")`. However, this application can analyze only parallel group designs with and without repeated measurements or factorial designs without interactions. The user receives a report including input files, the R markdown file, and R scripts necessary to replicate the analysis using their in-house R installation. In this way, `r BiocStyle::Githubpkg("fgcz/prolfqua")` and [B-Fabric](https://fgcz-bfabric.uzh.ch/) and helps scientists to meet requirements from funding agencies, journals, and academic institutions while publishing their data according to the FAIR  [@FAIR] data principles. We are working on creating a shiny stand-alone application with the described functionality and making it available soon.

@@

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes receiver operator curves (ROC) and computes partial areas under those curves ($pAUC$) and other scores, e.g., the false discovery proportion $FDP$. We use those scores, i.e. the $pAUC$ at $10%$ $FDR$ and the $FDP$, to examine how well the methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` model quantitative mass spectrometric high throughput data and compare them with the results produced by `r Biocpkg("MSstats")`, `r Biocpkg("proDA")` and `r Biocpkg("msqrob2")`. Table \@ref(tab:describeBenchmarkedModels) summarizes all the modelling methods we evaluated.

@@

pB_pAUC <- ggplot2::ggplot(resAllB, ggplot2::aes(x = Name, y = pAUC_10)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::facet_wrap(~whatfix)  + 
  ggplot2::coord_cartesian(ylim = c(min(resAllB$pAUC_10),max(resAllB$pAUC_10))) + 
  ggplot2::theme_minimal() + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5)) +
  ggplot2::geom_hline(ggplot2::aes(yintercept = meanpAUC_10), color = "red") +
  ggplot2::xlab("") +
  ggplot2::labs(tag = "B")

@@

When comparing differential expression analysis performance, a relevant parameter is the number of proteins for which a method estimates differences (see Figure \@ref(fig:figure3) Panel A), which indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4046 proteins with at least two peptides, there are in total $16184$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as the $pAUC$. For instance, we observe that `r Biocpkg("MSstats")` estimates $16058$ group differences while the mixed effect models estimates the fewest with $15940$. Hence, to conclude that one method shows a better performance, it does not suffice if the $pAUC$ is greater, but the number of proteins with differential expression results needs to be equal or larger.

Figure \@ref(fig:figure3) Panel B shows how various estimates obtained from the models, i.e., the difference between groups, $t$-statistics and scaled $p$-values, allow identifying true positives (TP) given a false positive rate (FPR) of $10%$ by displaying the partial area under the ROC. Ordering proteins by the $t$-statistic or $p$-value leads to a higher $pAUC_{10}$ than when ordering by the estimated difference among groups. Suppose an accurate estimate of the difference among groups is essential. In that case, the linear models fitted to protein intensities, calculated using Tukey's median polish, perform better than those when directly modeling peptide intensities, e.g., `ropeca` or `prot_mixed_effect_moderated` (see Table \@ref(tab:describeBenchmarkedModels) Abundance column). We speculate that outliers at the peptide level do not affect the protein estimates when using Tukey's median polish method.

We also benchmark if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:figure3) Panel C shows the $FDP$, obtained from the confusion matrix, as a function of the $FDR$ determined from the model. Most lines are below the diagonal, which indicates that the $FDR$ estimates are modestly conservative for this particular benchmark dataset. In the case of `r Biocpkg("MSstats")`, we observe a high proportion of false discoveries for small $FDR$ values. In the case of the ROPECA method, the $FDR$ estimates, obtained by applying the Benjamini-Hochberg correction to the Beta distribution-based regulation probabilities, strongly overestimate the $FDP$.

@@

nrgg <- dds |> ggplot2::ggplot(ggplot2::aes(x = name, y = nrcontrasts )) + 
  ggplot2::geom_bar(stat = "identity", fill = "white", colour = "black") + 
  ggplot2::coord_cartesian(ylim = c(min(dds$nrcontrasts) - 200, max(dds$nrcontrasts) + 10)) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5)) +
  ggplot2::geom_text(ggplot2::aes(label = paste0(round(nrcontrasts, digits = 1), paste0("  (",round(percent, digits = 1),"%)"))),
            vjust = 0, hjust = -0.2, angle = -90) #+ 
pA_barplot <- nrgg + ggplot2::labs(tag = "A")

@@

pC_FDR <- ddb |> ggplot2::ggplot(ggplot2::aes(y = FDP,  x  = FDR )) + 
  ggplot2::geom_line(ggplot2::aes(color = model_name, linetype = model_name)) +
  ggplot2::facet_wrap(~contrast) + 
  ggplot2::geom_abline(intercept = 0, slope = 1, color = 2) + 
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(tag = "C") + ggplot2::xlim(c(0,0.5)) + ggplot2::ylim(c(0,0.5))
hlay = rbind(c(2,1,1),
            c(2,3,3))

gridExtra::grid.arrange(pB_pAUC, pA_barplot, pC_FDR, layout_matrix = hlay)

@@

Using a benchmark dataset whose ground truth is known (see Methods), we assessed the performance of different modeling approaches implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` (Tables \@ref(tab:prolfquaModels) and \@ref(tab:describeBenchmarkedModels)), `r Biocpkg("MSstats")`, `r Biocpkg("proDA")` and `r Biocpkg("msqrob2")`. Table \@ref(tab:describeBenchmarkedModels) summarizes which methods we have evaluated, which MaxQuant files we used as input, and if the models are fitted to peptide or protein intensities. We make the R-markdown files to replicate the benchmarking available at `r BiocStyle::Githubpkg("wolski/prolfquabenchmark")`.
